---
title: "Introduction to R and RStudio"
output: html_notebook
---

Welcome to R! This script introduces the building blocks of coding in R: creating variables, understanding data types, and using the pipe operator to build data pipelines.


## Part 1: Variables

A **variable** stores a value so you can refer to it later. In R, you create a variable using the assignment operator `<-`. Think of it as an arrow pointing a name at a value.

```{r}
# A numeric variable
rank <- 17

# A character (text) variable — always use quotes
song_title <- "Baile Inolvidable"

# A logical variable — only TRUE or FALSE
is_number_one <- FALSE
```

Type the name of any variable and run the line to see its value:

```{r}
rank
song_title
is_number_one
```

You can do math with numeric variables:

```{r}
peak_position <- 2
chart_fall <- peak_position - rank
chart_fall
```

A few things to know:
- Variable names cannot start with a number or contain spaces. Use underscores: `my_variable`, not `my variable`.
- R is case-sensitive: `Song` and `song` are different variables.
- Comments inside of code chunks start with `#` — R ignores everything after the `#` on that line.

---

## Part 2: Vectors

A **vector** is a collection of values of the same type. Create one with `c()`, which stands for "combine":

```{r}
# A numeric vector of chart positions
chart_positions <- c(1, 2, 3, 4, 5)

# A character vector of performer names
performers <- c("Ella Langley", "Olivia Dean", "Alex Warren", "Taylor Swift","Bruno Mars")
```

R has several built-in functions that work on vectors:

```{r}
length(chart_positions)   # how many values are in the vector
sum(chart_positions)      # add them all up
mean(chart_positions)     # average
min(chart_positions)      # lowest value
max(chart_positions)      # highest value
```

---

## Part 3: Reading Data

Real analysis starts with loading a dataset. The `read_csv()` function reads a comma-separated file into R and stores it as a **data frame** — a table with rows and columns.

Load the tidyverse, a collection of packages designed for data analysis:

```{r}
library(tidyverse)
```


We'll use Billboard Hot 100 chart data from the `data` folder:

```{r}
hot_100 <- read_csv("data/hot-100-current.csv")
```

The name on the left of `<-` is up to you. Pick something short and descriptive.

---

## Part 4: Exploring Your Data

Before you analyze, get to know your data. These functions give you a quick overview:

```{r}
# A compact summary of all columns and their data types
glimpse(hot_100)
```

The `glimpse()` output shows each column name, its type (`<chr>` = character, `<dbl>` = numeric, `<date>` = date), and a preview of the first few values.

```{r}
# Show the first 6 rows
head(hot_100)
```

```{r}
# Count rows and columns
nrow(hot_100)
ncol(hot_100)
```

```{r}
# List all column names
names(hot_100)
```

The hot_100 data frame has these columns:
- `chart_week` — date of the chart
- `current_week` — chart position this week (1 = #1)
- `title` — song title
- `performer` — artist name
- `last_week` — chart position last week
- `peak_pos` — highest chart position ever reached
- `wks_on_chart` — total weeks on the chart

---

## Part 5: The Pipe Operator

The pipe `%>%` is one of the most important tools in the tidyverse. It takes the output of one function and passes it as the input to the next function. Read it as **"and then"**.

**Keyboard shortcut:** On Mac, press `Cmd + Shift + M` to type `%>%`. On Windows/Linux, use `Ctrl + Shift + M`.

With the pipe, you write left to right in the order things happen:

```{r}
# With the pipe, getting the first 10 rows
hot_100 %>% head(10)
```

The real power of the pipe is chaining multiple steps into a readable pipeline:

```{r}
# Take hot_100, then pass it to head() for the first 10 rows
hot_100 %>%
  filter(chart_week == as.Date("2026-02-14")) %>% 
  arrange(desc(wks_on_chart)) %>% 
  select(current_week, title, performer, wks_on_chart)
```

---


## Practice

Try answering these questions by writing code in the chunks below.

**1. Create a variable called `my_name` that stores your first name as a character:**

```{r}

```

**2. Create a numeric vector called `ages` containing the five numbers of the ages of people around you, then find the mean:**

```{r}

```

**3. Use `glimpse()` on `hot_100` — how many rows does the dataset have?**

```{r}

```

**4. Use `select()` and the pipe to show only the `title`, `performer`, and `peak_pos` columns from `hot_100`:**

```{r}

```
