---
title: "Importing"
output:
  html_document:
    df_print: paged
---

As a reminder, always add your packages at the top of your script. We'll use the `readxl` package to import an .xlsx file, and `janitor` to clean up some column names.

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(googlesheets4)
```

### Importing flat text files

```{r}
ledger <- read_csv("data/ledger.csv")
```

### Importing csv files from Github

Navigate to a dataset (in this case, *ledger.csv*) on Github. Remember to click on the "Raw" button, which will open a new page in your browser. Copy that link!

```{r}
ledger_online <- read_csv("https://raw.githubusercontent.com/ireapps/NICAR26-Intro-to-R/refs/heads/main/data/ledger.csv")
```

### Importing excel files

This is relatively straightforward, and primarily, we use the function `read_excel()` to import .xlsx or .xls files. This function has very similar arguments to `read_csv()`, and you can use the col_names and col_types arguments in the same way. With `read_excel()`, you can also specify what sheet in an excel workbook you want to import. In our example, `salaries.xlsx`, the first sheet is the data and the second is a notes tab, so we'll just import the first: 
```{r}
boone_detainees <- read_excel("data/boone_detainees.xlsx")

# this loads the first sheet by default

# you can import other sheets (which are not the first one), by adding a sheet argument

# for example,

boone_detainees <- read_excel("data/boone_detainees.xlsx", sheet = 2) # this would import the second sheet in theory! In practice, this particular dataset only has one sheet, so this code will not work
```


### Importing directly from Google Sheets

The package `googlesheets4` lets you connect one or more google accounts with RStudio and import data directly from a google sheet. You can import the whole sheet or a particular range. The import functions in `googlesheets4` also take most of the arguments used in `read_csv()`, such as `col_names` and `col_types`. 

To import a google sheet, you need to authenticate an account (which the package will prompt you to do in the console once you initiate an import) and you need the ID for the sheet, which is the long string of characters in the URL after *https://docs.google.com/spreadsheets/d/*. The string you want ends right *before* */edit*.

For this dataset of mlb salaries (source: https://docs.google.com/spreadsheets/d/1YctIvE6PBjYujA-Y-bPUcIhF8C1BxUyD4woG2Loskdo/edit?gid=0#gid=0) you're about to import, the string is "1YX1Z5FEbTdgFSPjFPmqbZkxIf-PUmUPtIo4-29dbP2k".

Use the function `read_sheet()`:
```{r}
grants <- read_sheet("1YctIvE6PBjYujA-Y-bPUcIhF8C1BxUyD4woG2Loskdo")

# this function has both sheet and range arguments. You can identify a particular sheet (think tab in a spreadsheet) to import by using that sheet's name: 

grants <- read_sheet("1YctIvE6PBjYujA-Y-bPUcIhF8C1BxUyD4woG2Loskdo", sheet="data")

# and you can import a particular range rather than the entire sheet:

grants <- read_sheet("1YctIvE6PBjYujA-Y-bPUcIhF8C1BxUyD4woG2Loskdo", range="A1:D953")

```

Note that you can import many different types of files to R, using several different packages. These are just the most commonly used by journalists. You just have to find the right function and the right package for what you need. Check out the `foreign` package, for example, which is great at helping you import large survey files of various types. 

# Tweaking imports

The following will apply to different files as you import them into R. We will be using the same ledger dataset for this. I am going to save each option as a separate dataset, so you can see what exactly happens with each one.

As a reminder, this is how our data looks when we don't add any tweaks. 

```{r}
ledger <- read_csv("data/ledger.csv")
```

Note that this file has a header row, although some of the names aren't very tidy (two have a space, and one has an odd character, a `#`. R will allow these to be column names but you need to put them between ` ` marks for R to recognize them as such).

If we use the `col_names` argument to say that this file does not have a header row, it will regard the first row as a row of data:

```{r}
ledger_no_header <- read_csv("data/ledger.csv", col_names=FALSE)
```

If we use `col_names` to specify a new set of column names, it will also assume the file does not have a header row:

```{r}
ledger_our_names <- read_csv("data/ledger.csv", col_names=c("col1", "col2", "col3", "col4", "col5", "col6"))

ledger_our_names
```


Let's talk data types. There's one column in here I might want to change: the Check # column. It's not a true number (I wouldn't want to do math on it), so I might want to import it as a character field. Use `col_types` to specify one character type for each column. See the documentation for the different data types:

-   c = character
-   i = integer
-   n = number
-   d = double
-   l = logical
-   f = factor
-   D = date
-   T = date time
-   t = time
-   ? = guess
-   \_ or - = skip

For this file, I want four character columns, one date time column and one number column, in that order:
```{r}
ledger_exact_col_types <- read_csv("data/ledger.csv", col_types="ccccTn")

ledger_exact_col_types
```

Alternately, I could import everthing as character. Sometimes this is the easiest way to import really dirty data, and then you can clean things up and convert columns to dates and numbers as you see fit: 
```{r}
ledger_all_c <- read_csv("data/ledger.csv", col_types = cols(.default = "c"))

ledger_all_c
```

To change just one column type, enter `?` for the rest of the columns (this tells R to guess, which it's good at). Let's say we only want to make sure that the check # column is stores as a character:
```{r}
ledger_just_check <- read_csv("data/ledger.csv", col_types = "???c??")

ledger_just_check
```

Or use `cols` if you don't want to type a bunch of question marks:
```{r}
ledger_just_check_2 <- read_csv("data/ledger.csv", col_types = cols(`Check #` = "c"))

ledger_just_check_2
```


Finally, to clean up column names (i.e. remove spaces and weird characters), use `clean_names()` from the `janitor` package:
```{r}
ledger_clean_names <- ledger %>% clean_names()

ledger_clean_names
```


You can also explore the options associated with `read_csv()` by looking at the docs.

```{r}
?read_csv
```

Note that there are a group of functions that show up, including read_tsv and read_delim. These are functions you can use for files with different formatting: tsv imports a tab-delimited file, and read_delim allows you to specify another delimiter.
Explore the documentation to see how you can further customize your imports. For example, you can skip rows if your file has extra header rows with the `skip = 2` argument. 

# Exporting a dataset

```{r}
write_csv(ledger_clean_names, "data/ledger_clean_names.csv")
```





